Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-18 manuscript/paper.tex
bib file added... some citations added to tex.

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL2010,CammarataJCP2014,nick2016,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL2010}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP2014}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{nick2016}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

@@

Since the octahedra are connected to each other 
(making a 3-dimensional network in the perovskite structure),
rotating one octahedron rotates all other in the crystal in a consistent fashion.
We consider an octahedral \emph{rotation} $\beta$ around $001$ direction
(that transforms the structure from cubic Pm$\bar 3$m to tetragonal P4/mbm)
%({No.~$127$})
followed by a \emph{tilting} $\alpha$ around the new $110$ direction
(transforming to orthorhombic Pbnm, 
%({No.~$62$})
or Imma 
%({No.~$74$})
 at ${\beta=0}$).
We use the rotation and tilt terminology for these 
two angles throughout the rest of the manuscript.

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B.
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
the splittings in its $t_{2g}$ and $e_g$ manifolds
that are
created primarily by the anion's octahedral potential.
We find that the size of the splitting 
and local orbital ordering
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-1}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-cubic}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-pbnm-15-15}\\
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-cubic-z}
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-rotated-15-15-z}
%\includegraphics[height=6.cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/AB-cages-deformed}
\caption{
Crystal structures and coordination cages.
%Crystal structure of cubic and orthorhombic perovskites (a-b) and ion coordination cages around B ions (c-d).
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:cages}
\end{center}
\end{figure}

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~(see \ref{append:multipoles}), 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we only have d-orbitals ($l=2$) on the central atom,
the matrix elements of its crystal field Hamiltonian
are given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the d-orbitals, assumed to be the same for all d-orbitals, i.e., the radial part of all d-orbitals is assumed the same~\cite{paxton-notes}.

@@

cubic ABO3 crystal Structure... 
rigid rotation/tilting: orthorhombic deformation... 
typically associated with the relative sizes of the A, B,O ions.~\cite{goodenoughPRL05}.

@@

It can be calculated using the method described in Ref.~\cite{evanic}.

@@

In a crystal, obviously, not only the nearest coordination cage of A,B ions but also
 the other ions (including O) at longer distances
  participate in the effect considered in this paper.
To see how much our results change when 
%the potential of 
 the full lattice (FL) is considered instead of just the nearest AB cages,
we numerically calculate the multipole components of the potential
 of the full lattice by following the recipe in Ref.~\cite{PaxtonNotes, book?}
and Ewald summation method~\cite{Ewald}.
To do this, 
we borrow relevant routines from the Tight Binding code TBE [confirm the names] in Questaal~\cite{https://www.questaal.org/} that implements these methods.

@@

>>>>

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL10,CammarataJCP14,LeeCM16,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL10}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP14}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{LeeCM16}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-1}
\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-2}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-cubic}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-pbnm-15-15}\\
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-cubic-z}
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-rotated-15-15-z}
%\includegraphics[height=6.cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/AB-cages-deformed}
\caption{
Crystal structures and coordination cages.
%Crystal structure of cubic and orthorhombic perovskites (a-b) and ion coordination cages around B ions (c-d).
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:struc}
\end{center}
\end{figure}

@@

Since the octahedra are connected to each other 
(making a 3-dimensional network in the perovskite structure),
rotating one octahedron rotates all other in the crystal in a consistent fashion;
 see Figs.~\ref{fig:struc}(a,b).
We consider an octahedral \emph{rotation} $\beta$ around $001$ direction
(that transforms the structure from cubic Pm$\bar 3$m to tetragonal P4/mbm)
%({No.~$127$})
followed by a \emph{tilting} $\alpha$ around the new $110$ direction
(transforming to orthorhombic Pbnm, 
%({No.~$62$})
or Imma 
%({No.~$74$})
 at ${\beta=0}$).
We use the rotation and tilt terminology for these 
two angles throughout the rest of the manuscript.

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B, as shown in Figs.~\ref{fig:struc}(c,d).
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
the splittings in its $t_{2g}$ and $e_g$ manifolds
that are
created primarily by the anion's octahedral potential.
We find that the size of the splitting 
and local orbital ordering
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

@@

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~(see \ref{append:multipoles}), 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we only have d-orbitals ($l=2$) on the central atom,
the matrix elements of its crystal field Hamiltonian
are given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the d-orbitals, assumed to be the same for all d-orbitals, i.e., the radial part of all d-orbitals is assumed the same~\cite{PaxtonNotes}.

@@

cubic ABO3 crystal Structure... 
rigid rotation/tilting: orthorhombic deformation... 
typically associated with the relative sizes of the A, B,O ions.~\cite{ZhouPRL05}.

@@

It can be calculated using the method described in Ref.~\cite{IvanicJPC96}.

@@

In a crystal, obviously, not only the nearest coordination cage of A,B ions but also
 the other ions (including O) at longer distances
  participate in the effect considered in this paper.
To see how much our results change when 
%the potential of 
 the full lattice (FL) is considered instead of just the nearest AB cages,
we numerically calculate the multipole components of the potential
 of the full lattice by following the recipe in Ref.~\cite{FinnisPRL98,PaxtonPRB08,PaxtonNotes}
and Ewald summation method~\cite{Ewald}.
To do this, 
we borrow relevant routines from the tight binding package in Questaal suite~\cite{Questaal} that implements these methods.

@@

\label{Bibliography}
\bibliographystyle{apsrev4-1}
\bibliography{perovskites}
